---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Skeena Moose Survey Stratification

<!-- badges: start -->
[![Lifecycle:Experimental](https://img.shields.io/badge/Lifecycle-Experimental-339999)](<Redirect-URL>)
<!-- badges: end -->

[SKmoose](https://gcperk.github.io/skmoose/index.html) is used to assess moose habitat quality within given survey blocks under aseries of criteria. This package provides a number of functions to extract data, estimate habitat and inhabitable areas within each block and report the values in a table. 

Where possible, data used in the process is extracted directly from the BC data catalogue, using [bcdata](https://github.com/bcgov/bcdata) and [bcmaps](https://github.com/bcgov/bcmaps) R packages to ensure up to date information is used. 

This package was written for Skeena Region under the Ministry of Water, Lands & Resources Stewardship. 

Optional additional data can be extracted including confidential survey telemetry data, however this is dependant on outside sources. 

A vignette is provided as a guide for users. These include a detailed step by step of all function in the package resulting in a csv table with compiled results. A test dataset is provided within the package to assist with understanding the process. 


## Installation

You can install the development version of skmoose from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("gcperk/skmoose")
```

## Overview

Potential Moose habitat is estimated for each survey block based on a series of criteria with spatial data extracted from the bcdata catalogue. Uninhabitable areas are also estimated using a set of criteria. All spatial data is extracted on a per block area rather than an entire study area to improve efficiency of processing. Appendix A provides a full list of references and spatial queries.
Once potenital habitat and inhabitable areas are compiled, total area of each are calculated to provide a means of stratifying moose survey blocks into catergories of Low, Medium and High. A table of all calculations is output for the user to review. 


## Criteria for potential Moose habiat

The following criteria are used to estimate area of moose habitat. Once compiled these areas are dissolved into a single polygon. Uninhabitable areas, as calculated below are then removed before final area calculations are completed. Values shown below in **bold** are the default values, which can be adjusted using parameters within the functions. 

- Deciduous tree species within the VRI (vegetation resource inventory). The default species include AC (Poplar), AT (Trembling Aspen), EP (Paper Birch), SB (Black Spruce), ie (**c("AT", "AC","EP","SB")**). 

- Early seral/shrub dynamic habitats. This includes: 1) fires and 2) cutblocks with a minimum time of **5 yrs** since disturbance and maximum time of **40 yrs** since disturbance. These ages can be adjusted seperately for each disturbance type. 

- Proximity to streams, streams with orders of **3 to 8** are buffered by **150m**, and streams with order of **9** are buffered to **500m**.

- Small lakes and wetlands which are less than 1 km2. 

- Skeena Wildlife Ecological Resource Model- Winter forage output - still to be added. 



## Criteria for uninhabitable areas 

Uninhabitable areas within the landscape are defined under the following criteria 

- Rock and Ice 

- Large waterbodies greater than 1km2

- High elevation areas. The default cutoff is currently set to areas above **1500 m** 

- Steep slopes. The default is currently set to greater than **55 degrees**. 


## Calculate proportion of Moose habitat

Once the area of moose habitat and uninhabitable areas has been calculated, we then estimate the area and combine these into a single table. 

The table contains the following fields: 

- bk = block id number

- total_area_m2 = total area of block in m2

- uninh_area_m2 = total uninhabitable area of the block (m2)

- hab_area_m2 = total habitable area within each block (m2)

- prop_uninh_block = proportion of uninhabitable area of each block. (uninh_area_m2/total_area_m2)

- net_habitat_area_m2 = Net area of habitat available within each block (m2) (total_area_m2 - uninh_area_m2) 

- prop_habit_block_m2 = Proportion of habitat within the total area of the block (m2) (hab_area_m2/total_area_m2) 

- prop_habit_net_habit_m2 = proportion of habitat within the net area of the block (hab_area_m2/net_habitat_area_m2)
    

### Appendix A 

The following datasets are used to estimate habitat and inhabitable areas. These include: 

- [VRI](https://catalogue.data.gov.bc.ca/dataset/vri-2022-forest-vegetation-composite-polygon). Vegetation Resource Inventory: Rock and Ice (uninhabitable) is filtered using the British Columbia Land Cover Classification Scheme Level 3 where values are classed as alpine. (BCLCS_LEVEL_3 = “A”)

- [VRI](https://catalogue.data.gov.bc.ca/dataset/vri-2022-forest-vegetation-composite-polygon). Vegetation Resource Inventory: Deciduous tree species (habitat) are filtered using the following fields: "SPECIES_CD_1", "SPECIES_CD_2","SPECIES_CD_3","SPECIES_CD_4","SPECIES_CD_5","SPECIES_CD_6". The default species of interest include (**c("AT", "AC","EP","SB")**). 

- [Lakes](https://catalogue.data.gov.bc.ca/dataset/freshwater-atlas-lakes) are filtered by area (AREA_HA) is greater than 1km2 (uninhabitable areas) and also < 1km2 for habitable area. 

- [wetlands](https://catalogue.data.gov.bc.ca/dataset/freshwater-atlas-wetlands) of area < 1kms are combined with small lakes as potential moose habitat.

- [stream network](https://catalogue.data.gov.bc.ca/dataset/92344413-8035-4c08-b996-65a9b3f62fca) is used to determine and buffer based on stream order. 

- [fires historic](https://catalogue.data.gov.bc.ca/dataset/22c7cb44-1463-48f7-8e47-88857f207702) and [fires current](https://catalogue.data.gov.bc.ca/dataset/cdfc2d7b-c046-4bf0-90ac-4897232619e1) are used to estimate fires. Time since burn is used to filter the miniumum and maximum range to include for moose habitat. 

- [cutblocks](https://catalogue.data.gov.bc.ca/dataset/b1b647a6-f271-42e0-9cd0-89ec24bce9f7) are filterd usein minimum and maximum time since harvest.

- [digital elevation model (dem)](https://rdrr.io/github/bcgov/bcmaps/man/cded.html) is used to estimate elevation and slope. This is extracted from the bcmaps packages and uses TRIM DEM. 

```{r, echo = FALSE}

# You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

```


